---
title: "Statistical Measures for Species Diversity"
output: 
  learnr::tutorial:
    css: css/style.css
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)

library(dplyr)
library(ggplot2)
library(ggfortify)
library(learnr)
library(knitr)
library(DT)
library(shiny)
library(gt)
library(plotly)
library(tidyverse)
library(vegan)
library(png)
library(jpeg)
library(grid)


plot_theme<-theme_light()+
  theme(plot.title=element_text(face =2,size=18),
        axis.title=element_text(face =2,size=14),
        plot.subtitle=element_text(face =3,size=14),
        axis.text.x=element_text(face =1,size=12),
                axis.text.y=element_text(face =1,size=7),
        strip.text=element_text(face =1,size=12))

datafiles<-readRDS("data/data.RDS")

scotland=datafiles$scotland
birds_2024=datafiles$birds_2024
bird_data=datafiles$bird_data
study_56=datafiles$study_56

```

## Introduction

In this module we are going to start looking into statistical techniques
which can help us to explain species diversity across different
environments.

-   Unpacking the different components of diversity analysis: richness,
    abundance, dominance and evenness
-   Reviewing some of the common indices used to bring together these
    measurements - the Shannon and Simpson indices
-   Highlighting important caveats and limitations around these measures
    of diversity, and introducing some strategies for where to go next
    after an initial analysis made through these measures.

The terminology used is often a little abstract, and often feels a
little unconnected to other areas of statistics, and the caveats around
the usage of these measures are extremely important to recognise when
trying to interpret measures of diversity.

I talk through in the video here:

## Data Used

The data is taken from the 2024 RSPB "Big Garden Birdwatch" - a citizen
science project encouraging anyone in the country to record and take
photos of which birds they observed in the area around their homes on a
particular day. This data is summarised at county level and provides the
average number per household for each of 80 different species that were
recorded.

The data is made up of reports provided by 610,000 households in the UK,
and a total of 9.5 million birds were reported by those households.

```{r cars}
DT::datatable(birds_2024)
```

We do have to be a bit careful with this data as a result - since there
may be some potential bias due to the way in which it was collected. In
each county there would have been different numbers of people reporting,
and the numbers reporting may not have been proportional to the
population sizes, or evenly distributed. There would also be questions
over whether households were able to accurately identify each bird,
particularly for those less common species.

## Components of Diversity

Within any data which has counted observed frequencies of different
species there are two easy ways in which we can start to assess the
diversity within each observed are:

a)  how many different species were reported
b)  how many total species were reported, summing across all of the
    different species

These two concepts, while simple, are fundamental to statistical ecology
and are referred to as the "Species Richness", the total number of
**different** species within the area of interest, and the "Total
Abundance", the total number of species present in the given area.

The *Abundance* would be the total number of a given species within a
given area of interest. So in this data we may be thinking of this in
terms of, for example:

-   "the *abundance* of Skylarks with Wiltshire"
-   "the *total abundance* of all birds within Wiltshire"
-   "the *species richness* of birds within Wiltshire"

### Richness

For the data we are exploring it is easy to get the assessment of the
species richness by identifying how many of the 80 species were
observed:

```{r}
diversity_data<-bird_data %>% 
  group_by(County) %>%
  summarise("Species Richness"=sum(Mean_2024>0),"Total Abundance"=sum(Mean_2024),
            "Dominance"=max(Mean_2024/sum(Mean_2024)),
            Shannon=vegan::diversity(as.numeric(Mean_2024),index="shannon"),
            Simpson=vegan::diversity(as.numeric(Mean_2024),index="simpson"),
            invsimpson=vegan::diversity(as.numeric(Mean_2024),index="invsimpson"),
            InvSd=1/sd(Mean_2024[Mean_2024>0]),
            PilouJ=Shannon/log(`Species Richness`))
```

```{r,fig.height=10}
(diversity_data %>%
  mutate(County=reorder(County,`Species Richness`))%>%
         ggplot(aes(x=`Species Richness`,y=County,fill=`Species Richness`))+
  geom_col()+
   plot_theme+
   theme(legend.position = "none")+
  ggtitle("Species Richness",subtitle = "Number of Different Species"))
```

Dumfries and Galloway had the most different number of bird species
identified, 52, and the Isles of Scilly had the fewest.

### Abundance

The abundance, and total abundance, are a little more challenging! We
are dealing with very large geographic areas (counties), with
non-randomly selected sampling units (households who chose to
participate), a three-day observation period and a count of species
which by definition do have a tendency to move around quite a lot
(birds).

So it would be impractical to try to calculate any sort of estimate for
the total number of birds per county based on the data collected. But
what we can estimate from the data is a "per-household" estimate of the
abundance of the different bird species - which will not impact on any
of the statistical measures in terms of calculation but only in terms of
interpretation. When dealing with these sort of diversity analyses it is
often necessary or advisable to be considering making adjustments like
this to enable for a better understanding of the site diversity.

Having a "per-household" estimate of species abundance is in itself not
so common outside of citizens science initiatives like this; instead it
is a convenient measure to give a proxy for the overall abundance. But
making adjustments of abundance and diversity per area would be
extremely common. This could also be something we could consider with
this dataset.

Another common adjustment would be to go beyond the "abundance", of
simply counting the species, and instead consider the biomass of our
species. This would be particularly important when considering an
ecosystem containing species of very different sizes:

For example: In Woodland A we observed 6 foxes, 20 rabbits and 100 ants
In Woodland B we observed 1 fox, 1 rabbit and 250 ants

The 'total abundance' in Woodland B is about double that in Woodland A;
but the "total biomass" in Woodland A is around 9 times larger than that
of Woodland B.

Given the more limited range of biomass between different bird species,
that sort of adjustment is unlikely to make too much of an impact here.

Let's compare how the total abundance (per household) and the species
richness compare to each other:

```{r}
(ggplot(diversity_data,aes(y=`Total Abundance`,x=`Species Richness`,County=County))+
  geom_point()+
   ylab("Average Abundance per Household")+
   xlab("Number of Different Species: Species Richness")+
  plot_theme) %>% ggplotly()
```

We can see there is, perhaps surprisingly, very little correlation
between the two measures across our counties. We already noted that
Dumfries and Galloway had the most different species of any county, and
this also had a high average number of birds observed per household. But
there are many counties with only a few less species identified that
have a much lower average abundance - Dorset, Somerset and Edinburgh for
example. And there are also counties with low numbers of different
species that have even higher average numbers of birds identified.

Hovering over those three points you may notice a common pattern: they
relate to Isles of Scilly, Shetland Islands and Eilean Siar (formerly
known as the Western Isles).

### Dominance

One aspect linked directly to the relative abundance of species that can
be useful to help our understanding of diversity is "dominance". The
"Berger-Parker" dominance score is one which is very simple to
calculate - it is simply the proportion of the total abundance that
comes from the single most dominant species.

Let's take a look at two of our more extreme counties Cornwall and the
Shetland Islands:

```{r,fig.height=6}
bird_data %>%
  mutate(County=trimws(County),Species=trimws(Species)) %>%
  filter(County%in%c("Cornwall","Shetland Islands")) %>%
  filter(as.numeric(Mean_2024)>0) %>%
  group_by(County) %>%
    mutate(RelativeAbundance=Mean_2024/sum(Mean_2024)) %>%
  ungroup() %>%
  mutate(Species=reorder(str_replace_all(Species,"_"," "),as.numeric(RelativeAbundance))) %>%
  ggplot(aes(y=Species,x=as.numeric(RelativeAbundance)))+
  geom_col(col="black")+
  plot_theme+
  xlab("Relative Abundance (% of birds in county)")+
  facet_wrap(~County,nrow=1)+
  scale_x_continuous(labels=scales::percent)
```

The dominance of the most common species in Cornwall, the House Sparrow,
is fairly low - it represents 16% of the total population of birds
observed. In the Shetland islands almost 50% of all birds observed were
sparrows.

This simple measure of dominance immediately gives us some insights into
the diversity of the two counties - but it is also a very limiting
statistic. It will, by definition, only focus on the most extreme
species.

In fact, with the Shetland Islands, were we to consider the two most
dominant species then we would see an even more extreme difference in
this metric. The top two species account for 75% of the total birds
observed in the Shetland Islands. So to try to generalise this idea more
widely, beyond the single extreme point or making an arbitrary choice
about how many "top" species to include in our measure of dominance,
then we consider the concept of "evenness".

### Evenness

The final component which contributes towards the measures of
diversity - evenness - tells us how evenly distributed each species is
within an environment. For example: In both County A and County B we
observed 10 different species, and an average total of 20 birds. In
County A - all 10 bird species were observed an average of 0.5 times per
household In County B - one bird species was observed an average of 19.1
times per household, and the other 9 were observed an average of 0.1
times per household.

County A has a much higher evenness than County B; despite both having
the same species richness and the same total abundance.

When considering evenness and comparing it across different environments
we are no longer thinking about "abundance" of each species in an
environment, but thinking about the "relative abundance" of each
species. That is what percentage of the total species in a given
environment belong to a given species.

A common measure of evenness is known as "Pilou's J" - where a value of
1 would indicate the species were all observed with equal frequency, and
the smaller the value the less even the distribution. We will come back
to the formula for this in the next section, but first let's see how
this measure relates to the species richness and total abundance:

```{r}
(ggplot(diversity_data,aes(y=PilouJ,x=`Species Richness`,County=County))+
  geom_point()+
   ylab("Evenness (Pilou's J)")+
   xlab("Number of Different Species: Species Richness")+
  plot_theme) %>% ggplotly()
```

```{r}
(ggplot(diversity_data,aes(y=PilouJ,x=`Total Abundance`,County=County))+
  geom_point()+
   ylab("Evenness (Pilou's J)")+
   xlab("Average Abundance per Household")+
  plot_theme) %>% ggplotly()
```

From these plots the first thing that catches our attention is the four
very different looking counties, with low richness, low evenness and a
high average count of birds per household. From this, without looking at
the data, we can infer that in these counties there is likely to be a
very small number of bird species observed a very large number of times
per household on average.

Ignoring the four island counties we can see a slight negative
correlation between evenness and richness - those counties with more
different species of birds tended to have less even spread in the
relative abundance of each bird.

Our relationship between evenness and abundance is much less strong; in
fact outside of our island counties there seems to be almost no
correlation between the two.

Let's pick out some of the more extreme counties, to see how this looks
in relation to the actual data itself:

```{r,fig.height=6,fig.width=8}
bird_data %>%
  mutate(County=trimws(County),Species=trimws(Species)) %>%
  filter(County%in%c("Isle of Man","Shetland Islands","Dumfries and Galloway")) %>%
  filter(as.numeric(Mean_2024)>0) %>%
  group_by(County) %>%
    mutate(RelativeAbundance=Mean_2024/sum(Mean_2024)) %>%
  ungroup() %>%
  mutate(Species=reorder(str_replace_all(Species,"_"," "),as.numeric(RelativeAbundance))) %>%
  ggplot(aes(y=Species,x=as.numeric(RelativeAbundance)))+
  geom_col(col="black")+
  plot_theme+
  xlab("Relative Abundance")+
  facet_wrap(~County,nrow=1)+
  scale_x_continuous(labels=scales::percent)
```

-   The relative abundance within the Shetland Islands is dominated by
    just two species

-   Both Dumfries and Galloway and the Isle of Man have fairly evenly
    spread distributions - although house sparrows are common there are
    many other species contributing substantially to the total. But
    Dumfries and Galloway has more different species, hence a higher
    richness, and most of the species observed in Dumfries and Galloway
    were rare, which is why it scored slightly less in evenness than the
    Isle of Man.

## Diversity Indices

Diversity indices are a way of summarising down the trade-off between
evenness and richness of species within a given area to a single number,
for ease of comparison and use in further analysis. This is known as
"alpha diversity".

There are two fairly commonly indices the Shannon-Weiner index and the
Simpson index, both of these look to provide a trade-off between the
richness and the evenness. Part of the reason why they are so commonly
used is because the formulae to calculate them are relatively simple:


```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/formula.png")
grid.raster(img)
```

In most situations they will be extremely highly correlated, and we can
see that is the case here:

```{r}
(ggplot(diversity_data,aes(y=Shannon,x=Simpson,County=County))+
  geom_point()+
  plot_theme) %>% ggplotly()
```

But there are important differences between, as they are not always
guaranteed to show the same results.

### Simpson Index

Simpson's index can be interpreted as the probability that any two
individuals randomly selected from within the population will come from
different Species. If there is just one species present in a population
by definition this will be 0. As the number of species increases the
maximum possible score for the Simpson index will become closer to 1.
And as the evenness of the distribution between species increases then
the score will tend closer to the maximum potential score for the index
given the number of species observed. e.g. if just 3 different species
are observed the maximum possible score of the Simpson index is 0.666,
which would be achieved when all three Species have equal relative
abundance.

### Shannon-Weiner Index

The larger the value of the Shannon-Weiner index, then the higher the
diversity that exists within the population. Although it is difficult to
give the number any direct meaning or interpretation that is
particularly useful (although some people do try) the general
interpretation would be that a value of less than 2 indicates low
diversity; more than 2.5 indicates moderate diversity, and a value of
more than 3 indicates high diversity.

Theoretically there is no upper limit on the Shannon Index - the largest
value for a given number of species identified would be the log of the
number of species identified, but it is extremely unlikely to find a
situation where it will be less than 1 or more than 4.

A general interpretation of the value of the index can be given that it
is equal to the log of Richness multiplied by the Evenness.

As such the value of evenness we used in the previous section, Pilou's
J, is calculated on this basis: the calculated is made by taking the
value Shannon Index and dividing it by the log of the number of species.

### Differences between the two methods

The Simpson method places slightly greater weight onto the evenness of
the species, and the Shannon-Weiner method will put slightly more
emphasis on the richness.

You can see this by comparing Merthyr Tydfil and North Lanarkshire -
both score identically on the Shannon-Weiner index, but differently on
the Simpson index. Merthyr Tydfil has a higher richness than North
Lanarkshire but a lower evenness. These cancel each other out on the
Shannon-Weiner index, but the weighting in favour of evenness gives a
higher value for the Simpson index for North Lanarkshire.

```{r}
diversity_data %>%
  select(County, `Species Richness`,PilouJ,Shannon,Simpson) %>%
  filter(County%in%c("Merthyr Tydfil","North Lanarkshire")) %>%
  gt() %>%
  fmt_number(3:5)
```

The reverse can be seen when comparing Devon and Merseyside - the
Shannon index score indicates higher diversity in Devon over Merseyside,
where the Simpson index gives them both the same score. And in this case
Devon has a higher richness score and a lower evenness score, and this
is sufficient with the weighting in favour of richness in the Shannon
index to give a higher score for Devon, whilst the two cancel each other
out in the Simpson index.

```{r}
diversity_data %>%
  select(County, `Species Richness`,PilouJ,Shannon,Simpson) %>%
  filter(County%in%c("Devon","Merseyside")) %>%
  gt() %>%
  fmt_number(3:5)
```

### Important Considerations & Limitations

### Sampling Design

In this case we have used an observational citizen science dataset to
try to assess diversity.

There are many potential risks involved with this approach:

-   Whether all participants have correctly identified the species,
    particularly important for the rarer species

-   The extent to which the same birds may have been counted on multiple
    occasions - which could happen within the same household; or the
    same bird being counted by multiple households. Given birds are not
    particularly prone to sticking to administrative boundaries, the
    same bird may even be counted across multiple counties in the
    survey!

-   The characterisation of "county" as the environment is also
    problematic for multiple reasons, which will explore in the next
    section. But fundamentally there is also no scientific reason why an
    administrative unit would be a good unit of comparison, compared to
    an environment classified based on ecological or geographic factors.
    It is a convenient one though, and also likely helps with the
    promotion and publicity of the survey, since is a very human level
    of aggregation!

-   The geographic spread, and number of households participating would
    be reliant on self-selection. It's likely there may be some highly
    engaged community groups participating in the survey, but these
    would largely all report quite similar findings which may lead to
    bias in the results, compared to other communities where few people
    may chose to participate.

But with those challenges also come benefits which could not be achieved
through any other method! The scale of the survey, covering the whole of
the UK, and the regularity in which it is conducted using the same
methodology every year, gives an incredibly rich data source that would
be impossible to achieve through other methods.

When designing your own studies for these purposes it is really
important to think through design aspects for exactly how you plan to
collect the data - adapting to the context of the environments you are
studying, and the behaviours of the species involved. Other potential
options could be:

-   Purely observational data, as we have seen in the bird survey

-   Analysing physical samples - e.g. counting insects within soil
    samples taken from transects; or classifying micro-organisms within
    water samples at different sampling locations

-   "Catch and release" methods for larger species, using traps, and
    potentially using electronic tags to then be able to monitor the
    patterns of the same individuals over time, and ensure there is no
    "double counting"

-   Setting "video traps" for even larger species still, where human
    observation may impact upon usual behaviours

-   All would have slightly different sets of pros and cons, and may be
    better adapted to certain contexts.

One aspect that was easy to define in this study was the *taxa*, the
genus which would be considered within scope for the survey. The
objective was to record all birds, which is a pretty straightforward
category to define the limits of. But if your focus is more related to
understanding the full diversity within *environments,* then you would
have to think carefully about which species would be included. If we
wanted to try to classify all possible macro-fauna, flora, insects,
mammals, birds, along anything else that may exist within our
environment, then we would be setting ourselves a pretty tough challenge
and one where the results of any analysis would be hard to fully grasp
compared to breaking down the species into more coherent groups.

### Sample Size & Rarefaction

One thing we haven't considered about the strangeness of the four island
counties is whether the results may be different to other counties not
because of reduced diversity in the bird populations but because fewer
people live there. This data comes solely from volunteer members of the
population recording the birds they see in their garden and then
choosing to report. We would expect fewer people to report results in
counties with small populations, and the remote island counties, as well
as being quite geographically distinct, are also those counties with the
lowest populations.

This is particularly important in diversity analysis since observed
species richness is extremely dependent on the sample size, length of
observation period and area covered by the sample. Within this survey
there may have been tens of thousands of households participating in
Cornwall, but only tens of households participating in the Scilly Isles.
When thinking about "richness" in the data we see most of the species
reported are 'rare', only reported a low number of times per county. The
lack of observations of rare birds in some of the counties does not
necessarily mean that they were not present at all - the more different
locations and more people observing over a longer period of time, across
a wider geographic area within the county increases the potential for
the rarer species to be identified.

With a small number of observations and a small observation period we
are almost certain to under-estimate the species richness, and thus also
underestimate the "true" values of our diversity indices.

In order to be able to make direct comparisons we would need to assume
that the theoretical maximum capacity for diversity is the same in each
of the environments we are comparing. The Simpson and Shannon indices
are not measures which lend themselves particularly well to being
adjusted directly "per-capita" or "per-area". Cornwall with an area of
3,500 sq-km would definitely seem to have a much higher capacity for
more different birds being observed as compared to the Isles of Scilly
with an area of 16 sq-km.

But there are more nuanced methods to allow for improvements to the
comparability between these measures through adjustments called
"rarefaction". You can read more about this here:
<https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2019.02407/full>
<https://www.scielo.sa.cr/scielo.php?script=sci_arttext&pid=S0034-77442009000300001>

Without the raw data at household level this is not something we can
really pursue too much further with this dataset.

### Other ways of considering diversity measurement

These measures of "alpha" diversity also fail to capture diversity
**across different environments**. Even where we do compare the
Shannon/Simpson index across environments they only compare the internal
levels diversity - they do not account for whether the same species are
being seen in different environments, whether there exist certain
species are seen only in small numbers of environments, or certain
environments with large numbers of rare species, or whether different
species are dominant in different environments.

These are all concepts which link to what is known as "beta" diversity,
which focuses on this diversity in the species population across
environments. There are a lot of different metrics that can be used to
understand this concept of "Beta Diversity" by comparing the similarity
in the populations of between sites.

There is an often cited paper that outlines 24 different options here:
<https://besjournals.onlinelibrary.wiley.com/doi/10.1046/j.1365-2656.2003.00710.x>

But that is only the starting point - as that paper solely considers
beta diversity in terms of presence/absence - there are many more than
that when starting to consider the difference in abundances between
sites! There are many methods for continuing further with this
investigation into diversity - techniques like cluster analysis or
network analysis can help to build a picture of how species may coexist
with each other, and which environments have similar ecological
profiles.

There is a good overview of the limitations of "alpha" diversity
measures, and some of the entry points to understanding beta diversity
here:
<https://ecologyforacrowdedplanet.wordpress.com/2016/01/14/beta-diversity-what-is-it-good-for/>

## Exercises

In these exercises we are going to look at data from a study looking at
the rodent populations near to Albuquerque, New Mexico.

Data is available from 1989 to 2008, with the same sampling processes
and locations used each year through a "trap and release" sampling
procedure. You can read more about the trial and the methods they used
here:

<https://biotime.st-andrews.ac.uk/selectStudy.php?study=56>

(And if you would like to look at some cute pictures of rodents to
distract you from statistics for a while - I highly recommend searching
for pictures of some of these species based on their name!)

```{r}
DT::datatable(study_56)
```
And you can download the raw data used here:

```{r dl, context="render",echo=FALSE}

downloadButton("downloadData", "Download Data")
```

```{r dl_server,context="server"}

output$downloadData <- downloadHandler(
    filename = function() {
      # Use the selected dataset as the suggested file name
      paste0("biotime_study56",".csv")
    },
    content = function(file) {
      # Write the dataset to the `file` that will be downloaded
  
      write.csv(study_56, file)
    }
  )

```



The general question we want to explore is:

*Has the diversity of the rodent population within the study area
changed over time? And if so how?*

Below you can explore the data in a few different ways below to help you
make this assessment, by looking at different visualisations of the data
over time and different diversity statistics.

Remind yourself of what each of the diversity statistics can tell you,
and what limitations they may have. See if you can link the diversity
statistics to the raw data.

And this dataset captured both the abundance and the biomass of the
rodents, you can choose to explore the diversity statistics and trends
based on either biomass or abundance.

*How does changing the choice of variable to assess diversity here
change the interpretation of diversity in the population?*

*Which statistics are most affected by the change of variable?*

#### Tables of Summary Statistics

```{r ex1_ui, context="render",echo=FALSE}

selectInput("calculations",label="Statisics based on:",choices=c("Abundance","Biomass"))
gt_output("table1")
```

```{r ex1_server, context="server"}
output$table1<-render_gt({
  if(input$calculations=="Abundance"){
    
    dominant<-study_56 %>%
      group_by(YEAR) %>%
      filter(ABUNDANCE==max(ABUNDANCE)) %>%
      mutate(`Dominant Species`=paste(GENUS,SPECIES)) %>%
      select(YEAR,`Dominant Species`)
        
  diversity_56<-study_56 %>% 
    inner_join(dominant)%>%
  group_by(YEAR,`Dominant Species`) %>%
   summarise("Species Richness"=sum(ABUNDANCE>0),"Total Abundance"=sum(ABUNDANCE),
            "Dominance"=max(ABUNDANCE/sum(ABUNDANCE)),
            "Shannon-Weiner"=vegan::diversity(as.numeric(ABUNDANCE),index="shannon"),
            "Simpson"=vegan::diversity(as.numeric(ABUNDANCE),index="simpson"),
            "Pilou-J Evenness"=`Shannon-Weiner`/log(`Species Richness`))%>%
    ungroup()%>%
    select(YEAR,`Species Richness`:Dominance,`Dominant Species`,`Shannon-Weiner`:`Pilou-J Evenness`)
  }
  else{
        
    dominant<-study_56 %>%
      group_by(YEAR) %>%
      filter(BIOMASS==max(BIOMASS)) %>%
      mutate(`Dominant Species`=paste(GENUS,SPECIES)) %>%
      select(YEAR,`Dominant Species`)
        
  diversity_56<-study_56 %>% 
    inner_join(dominant)%>%
  group_by(YEAR,`Dominant Species`) %>%
   summarise("Species Richness"=sum(BIOMASS>0),"Total Biomass"=sum(BIOMASS),
            "Dominance"=max(BIOMASS/sum(BIOMASS)),
            "Shannon-Weiner"=vegan::diversity(as.numeric(BIOMASS),index="shannon"),
            "Simpson"=vegan::diversity(as.numeric(BIOMASS),index="simpson"),
            "Pilou-J Evenness"=`Shannon-Weiner`/log(`Species Richness`)) %>%
    ungroup() %>%
    select(YEAR,`Species Richness`:Dominance,`Dominant Species`,`Shannon-Weiner`:`Pilou-J Evenness`)
    
  }
  gt(diversity_56) %>% fmt_number(c(4,6:8),decimals=3)

})
```

#### Within Year Species Counts
```{r ex_x_ui, context="render",echo=FALSE}

selectInput("calculationsx",label="Statisics based on:",choices=c("Abundance","Biomass"))
selectInput("year",label="Year to Plot",
            choices=1989:2008)
plotlyOutput("plotx",height = "600px")
```

```{r ex_x_server, context="server"}

output$plotx<-renderPlotly({
  study_56$var<-study_56[[toupper(input$calculationsx)]]
  (study_56 %>%
      filter(YEAR==as.numeric(input$year)) %>%
      ungroup() %>%
    mutate(genus=reorder(paste(GENUS,SPECIES),var,mean)) %>%
  ggplot(aes(y=genus,x=var))+
    geom_col(col="black")+
    xlab(paste("Total",input$calculationsx))+
      plot_theme+
    ggtitle(input$year)) %>% ggplotly()
  
  
})
  
```


#### Time Series of Summary Statistics

```{r ex2_ui, context="render",echo=FALSE}

selectInput("calculations1",label="Statisics based on:",choices=c("Abundance","Biomass"))
selectInput("variable1",label="Summary Variable to Plot",
            choices=c("Species Richness","Total Abundance/Biomass"="Total","Dominance","Shannon-Weiner",
                      "Simpson","Pilou-J Evenness"))
plotlyOutput("plot1")
```

```{r ex2_server,context="server"}

output$plot1<-renderPlotly({
  
if(input$calculations1=="Abundance"){
    
    dominant<-study_56 %>%
      group_by(YEAR) %>%
      filter(ABUNDANCE==max(ABUNDANCE)) %>%
      mutate(`Dominant Species`=paste(GENUS,SPECIES)) %>%
      select(YEAR,`Dominant Species`)
        
  diversity_56<-study_56 %>% 
    inner_join(dominant)%>%
  group_by(YEAR,`Dominant Species`) %>%
   summarise("Species Richness"=sum(ABUNDANCE>0),"Total"=sum(ABUNDANCE),
            "Dominance"=max(ABUNDANCE/sum(ABUNDANCE)),
            "Shannon-Weiner"=vegan::diversity(as.numeric(ABUNDANCE),index="shannon"),
            "Simpson"=vegan::diversity(as.numeric(ABUNDANCE),index="simpson"),
            "Pilou-J Evenness"=`Shannon-Weiner`/log(`Species Richness`))%>%
    ungroup()
  }
  else{
        
    dominant<-study_56 %>%
      group_by(YEAR) %>%
      filter(BIOMASS==max(BIOMASS)) %>%
      mutate(`Dominant Species`=paste(GENUS,SPECIES)) %>%
      select(YEAR,`Dominant Species`)
        
  diversity_56<-study_56 %>% 
    inner_join(dominant)%>%
  group_by(YEAR,`Dominant Species`) %>%
   summarise("Species Richness"=sum(BIOMASS>0),"Total"=sum(BIOMASS),
            "Dominance"=max(BIOMASS/sum(BIOMASS)),
            "Shannon-Weiner"=vegan::diversity(as.numeric(BIOMASS),index="shannon"),
            "Simpson"=vegan::diversity(as.numeric(BIOMASS),index="simpson"),
            "Pilou-J Evenness"=`Shannon-Weiner`/log(`Species Richness`)) %>%
    ungroup

  }
  print(colnames(diversity_56))
    print(input$variable1)
  diversity_56$var<-diversity_56[[input$variable1]]
(diversity_56 %>%
  ggplot(aes(x=YEAR,y=var))+
  ylab(input$variable1)+
  geom_point()+
  geom_line()+
  plot_theme+
  ggtitle(paste(input$variable1,"based on:",input$calculations1)))%>%
  ggplotly()


  })  
```

#### Time Series of Each Species

```{r ex3_ui, context="render",echo=FALSE}

selectInput("calculations2",label="Statisics based on:",choices=c("Abundance","Biomass"))
plotlyOutput("plot3")
```

```{r ex3_server,context="server"}

output$plot3<-renderPlotly({
  if(input$calculations2=="Abundance"){
  (study_56 %>%
      mutate(GenusSpecies=paste(GENUS,SPECIES)) %>%
        ggplot(aes(y=ABUNDANCE,x=YEAR,shape=GENUS,colour=GenusSpecies))+
          geom_point()+
            geom_line()+
            plot_theme)->p1
  }else{
    (study_56 %>%
      mutate(GenusSpecies=paste(GENUS,SPECIES)) %>%
        ggplot(aes(y=BIOMASS,x=YEAR,shape=GENUS,colour=GenusSpecies))+
          geom_point()+
            geom_line()+
            plot_theme)->p1  
  }
  
  p1%>% ggplotly()
  
})
```

## Resources

Choosing and using diversity indices: insights for ecological
applications from the German Biodiversity Exploratories
<https://pmc.ncbi.nlm.nih.gov/articles/PMC4224527/>

Direct overview of different components of diversity:
<https://www.flutterbys.com.au/stats/tut/tut13.2.html>

Beta Diversity:
<https://besjournals.onlinelibrary.wiley.com/doi/10.1046/j.1365-2656.2003.00710.x>
